---
title: "Ec 122 - HW4"
author: "Ben Juarez"
date: "10/23/2021"
output: pdf_document
---

# HGL 4.6
The fitted least squares line is $\hat{y}_i = b_1 + b_2x_i$.

## Part a
Let us algebraically show that the fitted line passes through the point of the means, $(\bar{x}, \bar{y})$.  Since we have that $b_1 = \bar{y} - b_2 \cdot \bar{x}$ for the least squares estimator for $\beta_1$, we can clearly see that $(\bar{x}, \bar{y})$ must be on the line because we can simply plug in for $x_i = \bar{x}$ as follows:
$$
\hat{y}_i = \bar{y} - b_2 \bar{x} + b_2x_i \implies \hat{y}_i = \bar{y}
$$

## Part b
Let us algebraically show that the average value of $\hat{y_i}$ equals the sample average of $y$.  In other words, we will show that $\bar{\hat{y}} = \bar{y}$ where $\bar{\hat{y}} = \sum \hat{y_i} / N$:
$$
\bar{\hat{y}} = \sum \hat{y_i} / N = (\frac{1}{N}) \sum (b_1 + b_2x_i) = (\frac{1}{N}) (\sum b_1 + \sum b_2x_i)
$$
$$
\implies \bar{\hat{y}} = (\frac{1}{N}) (N \cdot b_1 + b_2 \cdot \sum x_i) = (b_1 + b_2 \cdot \frac{\sum x_i}{N}) 
$$
We know $\bar{x} = \sum x_i / N$, so we can arrive at the following:
$$
\implies \bar{\hat{y}} = b_1 + b_2 \cdot \bar{x} = \bar{y}
$$

# HGL 4.14
```{r}
setwd("~/Desktop/R")
data1 = read.table("dat/cps4_small.dat")
colnames(data1) = c("wage", "educ", "exper", "hrswk", "married", "female", "metro", 
                    "midwest", "south", "west", "black", "asian")
```

## Part a
```{r}
hist1 = hist(data1$wage, breaks = 30, xlab = "WAGE", main = "WAGE Histogram")
hist2 = hist(log(data1$wage), breaks = 30, xlab = "ln(WAGE)", main = "ln(WAGE) Histogram")

```
It appears that the histogram for ln(WAGE) is more normally distributed since the histogram for wage is skewed right.

## Part b
```{r}
educ = data1$educ
wage = data1$wage

wage_lm = lm(wage ~ educ)
wage_llr = lm(log(wage) ~ educ)
summary(wage_lm)
summary(wage_llr)

mean(wage)
unname(coef(wage_lm)[2]/mean(wage) * 100)
unname(coef(wage_llr)[2] * 100)
```
For an additional year of education, the percentage increase in wages for the average worker (20.6 dollars/hr) is either about $9.6\%$ (linear model) or $9.04\%$ (log-linear model).

## Part c
```{r}
hist3 = hist(resid(wage_lm), breaks=30, xlab = "residuals (wage_lm)", main = 
               "Linear Regression Residuals Histogram")
hist3 = hist(resid(wage_llr), breaks=30, xlab = "residuals (wage_llr)", main = 
               "Log-Linear Regression Residuals Histogram")

library("tseries")
jarque.bera.test(resid(wage_lm))
jarque.bera.test(resid(wage_llr))
```
The histograms and the results from the Jarque Bera Test imply that the residuals from the log-linear model appear more compatible with normality, relative to the linear model.

## Part d
For the linear model, the $R^2$ is 0.175, and for the log-linear model, the $R^2$ is 0.1782.  Although these values are similar, since 0.1782 > 0.175, the log-linear model fits the data better.

## Part e
```{r}
plot(educ, resid(wage_lm), xlab="EDUC", ylab="residual")
plot(educ, resid(wage_llr), xlab="EDUC", ylab="residual")
```

We can see a pattern such that as EDUC increases, the residuals also increase in magnitude.  Also, the residuals are essentially positive in the first quarter of the range for EDUC which could tell us something about the effect of EDUC up to a certain point (about educ = 12) when the residuals begin to change.

## Part f
```{r}
pred_lm = coef(wage_lm)[1] + coef(wage_lm)[2] * 16
pred_llr = exp(coef(wage_llr)[1] + coef(wage_llr)[2] * 16)
unname(pred_lm)
unname(pred_llr)
data_16 = as.data.frame(filter(data1, educ == 16))
mean(data_16$wage)
```
So, the predictions for the wage of a worker with 16 yrs. of education are $\approx 25$ (linear model) and $\approx 21$ (log-linear model).  With the actual average wage of all workers with 16 yrs. of experience being 25.5, we can see that the linear model has a more accurate prediction.

## Part g
I would use the log-linear model because it has a better $R^2$ value while also having residuals that more normally distributed.  It is important to recognize that the linear model did have a more accurate prediction in the previous problem. 

# HGL 5.3

## Part a
Using Table 5.6:

### Part i
t-statistic $= \frac{b_1}{se(b_1)} = \frac{0.0091}{0.0190} = 0.479$

### Part ii
Std. Error = $se(b_2) = \frac{0.0276}{6.6086} = 0.00418$

### Part iii
Coefficient = $0.0002 \cdot (-6.9624) = -0.00139$ 

### Part iv
We know $R^2 = \frac{SSR}{SST} = 1 - \frac{SSE}{SST}$:
$$SSE=5.752896 \quad SST = (N-1)\cdot s^2_y = (1519-1)\cdot (0.0633)^2 = 6.082459$$
$$
\implies R^2 = 1 - \frac{5.75286}{6.082459} = 0.0542
$$

### Part v
$\hat{\sigma}^2 = \frac{SSE}{N-K} \implies \hat{\sigma} = \sqrt{\frac{SSE}{N-K}} = \sqrt{\frac{5.752896}{1519-4}} = 0.0616$

## Part b
We can interpret $b_2 = 0.0276$ as the increase in WALC (household's budget on alcohol) if ln(TOTEXP) increases by 1.  We can interpret $b_3 = -0.00139$ as the decrease in WALC (proportion of household's budget on alcohol) if AGE (age of household head) increases by 1 year.  We can interpret $b_4 = -0.0133$ as the decrease in WALC (proportion of household's budget on alcohol) if NK increases by 1 (household gains one more child).

## Part c
Let us compute a $95\%$ interval estimate for $\beta_3$ as follows:
$$
b_3 \pm t_{(1-\frac{\alpha}{2}, N-K)=(0.975,1515)} \cdot se(b_3) = -0.00139 \pm 1.9615 \cdot 0.0002 = (-.00178, -0.000998)
$$
This interval estimate tells us that if AGE (age of household head) increases by 1 year, then we are $95\%$ confident that WALC (proportion of household's budget on alcohol) decreases by a value in this range.

## Part d
Let us test the hypothesis ($5\%$ level of significance)that the budget proportion for alcohol does not depend on the number of children in the household: $H_0:B_0=0$ vs. $H_1:B_4\neq0$.  So, we know $t_c = t_{(0.975,1515)} = 1.9615$ and we reject if $t > t_c$ or $t < -t_c$.  We calculate $t$ as follows:
$$
t = \frac{b_4}{se(b_4)} = \frac{-0.0133}{0.0033} = -4.03
$$
Since $t=-4.03 < -t_c = 1.9615$, we reject $H_0$ and accept $H_1$.  This implies that NK (number of children in the household) does have an impact on the budget proportion for alcohol.  The reason for this test outcome could be that having one or more children takes up a lot of the household's budget, meaning there is less money to spend on alcohol.

# HGL 5.12
```{r}
setwd("~/Desktop/R")
data2 = read.table("dat/cocaine.dat")
colnames(data2) = c("price", "quant", "qual", "trend")
```

## Part a
For $B_2$, I would expect the sign to be negative because it seems to make that than with a larger sale of cocaine, there would be better prices per gram.  For $B_3$, I would expect the sign to be positive because it seems clear that higher quality cocaine would be more expensive.  For $B_4$, I would expect the sign to be positive because it might make sense that there would be more demand as time went on (unsure about supply).  I am not sure about the popularity of cocaine between 1984 and 1991, so I am unsure about this one.

## Part b
```{r}
PRICE = data2$price
QUANT = data2$quant
QUAL = data2$qual
TREND = data2$trend
coke_lm = lm(PRICE ~ QUANT + QUAL + TREND)
summary(coke_lm)
```
It appears that my expectations for $B_2$ and $B_3$ were correct, but I was wrong for $B_4$.  Perhaps demand did not match with increases in supply.

## Part c
Knowing that $R^2 = 0.5097$, we can say that $50.97\%$ of the variation in cocaine price is explained jointly by variation in quantity, quality, and time.

## Part d
Let us test $H_0:\beta_2 \geq 0$ vs $H_1:beta_2 < 0$.  We set up $H_0, H_1$ in this way because it makes sense that if sellers are not wiling to accept a lower price if they can make more sales in larger quantities, then $B_2$ would be positive.  Let us perform this test with $5\%$ significance as follows:
$$
t_c = t_{(0.95, 52)} = -1.675
$$
$$
t = \frac{b_2}{se(b_2)} = \frac{-0.05997}{0.01018} = -5.89 < t_c = -1.675
$$
So, we reject $H_0$ which means that it is true that sellers will accept a lower price if they can make sales in larger quantities.

## Part e
Let us test the hypothesis that the quality of cocaine has no influence on price ($\beta_3 = 0$) against the alternative that a premium is paid for better-quality cocaine ($\beta_3 > 0$): 
$$
H_0:\beta_3=0 \text{ vs } H_1:\beta_3 >0 \implies t_c = t_{(0.05, 52)} = 1.675
$$
$$
t = \frac{b_3}{se(b_3)} = \frac{0.1162}{0.20326} = 0.057 < t_c = 1.675
$$
So, we do not reject $H_0$, so we cannot conclude that a premium is paid for better-quality cocaine.

## Part f
We have that the average annual change in the cocaine price is -2.35.  This could potentially be explained by decreasing demand and/or oversupply.  There are a variety of factors that could play into these possible explanations.  Perhaps the risk of getting caught selling cocaine went down or perhaps those who frequently purchased cocaine suffered too much from the negative effects of the substance.
